Figures for Paper

### 2021-07-21 paper_figures.Rmd
Update paper_figures.Rmd

## Systematic differences in  (read length) between cases and controls

```{r sequences and case/controls}
# library size 
#llumina HiSeq 2500 (125 bp paired end) or an Illumina NovaSeq (100 bp paired end). 
case_control_sequencing <- clean_data_table %>%
  filter(tissue_clean %in% c("Frontal_Cortex", 
                             "Lumbar_Spinal_Cord", 
                             "Cervical_Spinal_Cord", 
                             "Motor_Cortex", 
                             "Temporal_Cortex",
                             "Cerebellum")) %>% 
  filter(sequencing_platform != "")

@@

    p_value = chisq.test(comparison_table) %>%  broom::tidy() %>% pull(p.value)
    browser()
    print(p_value)

@@

plot_list = list()
for(t in 1:length(lib_comps)){
    if(grepl("Cord",lib_comps[t]) | lib_comps[t] == "Motor_Cortex"){
        groups = c("Control","ALS-TDP","ALS \n non-TDP")
    }else if(lib_comps[t] %in% c("Frontal_Cortex","Temporal_Cortex")){
        groups = c("Control","FTLD-TDP","FTLD \n non-TDP")
    }else{
        groups = c("ALS-TDP", "Control", "ALS \n non-TDP", "FTLD-TDP", "FTLD \n non-TDP")
    }

@@

        ggplot(aes(x = disease_group2,y = rin,fill =disease_group2 )) + 
        geom_boxplot(colour = "black", outlier.colour = NA) +
        geom_point(pch = 21, size = 2, alpha = 0.9, position = position_jitter(width = 0.2, height = 0) ) + 
        facet_wrap(~ tissue_clean,scales = 'free_x') + 
        # ggpubr::stat_compare_means(label = "p.format", label.x.npc = "center", method = "wilcox" ) +
        ggpubr::stat_compare_means(label = "p.signif",
                                   label.x.npc = "center",
                                   method = "wilcox",
                                   comparisons = list(
                                       c("Control","ALS-TDP"),
                                       c("ALS \n non-TDP","ALS-TDP"))) +
        ggpubr::stat_compare_means(label = "p.signif", 
                                   label.x.npc = "center", 
                                   method = "wilcox",
                                   comparisons = list(
                                       c("Control","FTLD-TDP"),
                                       c("FTLD-TDP","FTLD \n non-TDP"))) + 
        labs(y = "RIN", x = element_blank()) +
        theme(axis.text.x = element_text(size = 15)) + 
        ggpubr::theme_pubr() + 
        theme(axis.ticks.x = element_blank(),
              axis.text.x = element_blank(),
              legend.position = "none")

@@

    if(grepl("Cord",lib_comps[t]) | lib_comps[t] == "Motor_Cortex"){
        groups = c("Control","ALS-TDP","ALS \n non-TDP")
    }else if(lib_comps[t] %in% c("Frontal_Cortex","Temporal_Cortex")){
        groups = c("Control","FTLD-TDP","FTLD \n non-TDP")
    }else{
        groups = c("ALS-TDP", "Control", "ALS \n non-TDP", "FTLD-TDP", "FTLD \n non-TDP")
    }

@@

cell_types %>% 
  left_join(clean_data_table[,.(sample,
                               UNC13A_3prime_leaf,
                               UNC13A_5prime_1_leaf,
                               disease_group2,
                               disease_tissue)], by = 'sample') %>%
  filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  filter(disease_tissue == T) %>% 
  filter(tissue_clean != "Thoracic_Spinal_Cord") %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-")) %>% 
  add_count(disease_group2,tissue_clean) %>% 
  mutate(n = n / 5) %>% 
  mutate(plot_name = glue::glue("{disease_group2} \n {n}")) %>% 
  filter(!is.na(inclusion_reads)) %>% 
  unique() %>% 
  ggplot(aes(x = cell, y = deconv, fill = unc13a_detected)) + 
  geom_point(size = 2,show.legend = F,pch = 21, position = position_jitterdodge(dodge.width=1,jitter.width = 0.3))+
  geom_boxplot(outlier.color = NA) + 
  facet_wrap(~plot_name+tissue,scales = 'free') +
  ggpubr::stat_compare_means(label = "p.signif") +
  ggpubr::stat_compare_means(label = "p.format",vjust = 1.2) +
  labs(title = "Cell type decomposition and UNC13A CE detection ",subtitle = "Wilcox test") + 
  ggpubr::theme_pubr()

@@

```{r onlyshort}
only_short_plotter = boxplot_table %>%
    filter(tissue_clean %in%
               c("Frontal Cortex",
                 "Lumbar Spinal Cord",
                 "Cervical Spinal Cord",
                 "Motor Cortex",
                 "Temporal Cortex",
                 "Cerebellum") )%>%
    group_by(disease_group2,tissue_clean) %>%
    mutate(n_sample = n_distinct(sample)) %>%
    ungroup() %>%
    mutate(disease_group2 = gsub("Control", " Control",disease_group2)) %>%
    mutate(detection_name = glue::glue("{disease_group2} \n ( {n_sample} )"))  %>%
    mutate(tissue_clean = fct_relevel(tissue_clean, "Cervical Spinal Cord",
                                      "Frontal Cortex",
                                      "Lumbar Spinal Cord",
                                      "Temporal Cortex",
                                      "Motor Cortex"))

comps = only_short_plotter %>% pull(detection_name) %>%
  unique() %>%
  combn(.,2,simplify = F)

only_short_plotter %>% 
    ggplot(aes(x = detection_name,
               y = inclusion_reads * 10^6,
               fill ="red")) +
    geom_boxplot(show.legend = F,position = position_dodge(preserve = "single",width = 1),outlier.colour = NA) +
    geom_point(size = 2,show.legend = F,pch = 21, position = position_jitterdodge(dodge.width=1,jitter.width = 0.3))+
    ggplot2::facet_wrap(vars(tissue_clean),scales = "free_x",nrow = 3) +
    ylab("UNC13A cryptic \n reads per million") +
    theme(text = element_text(size = 12)) +
    xlab("") +
    scale_fill_manual(values = colorblind_pal()(4)[2:4]) +
    theme_bw() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(text = element_text(size = 18))  +
    theme(legend.position = "bottom")
```

@@

## Side by side comparison UNC13A & STMN2 CE PSI
```{r}
clean_data_table %>%
    mutate(tissue_clean = gsub("_"," ",tissue_clean)) %>%
    filter(tissue_clean %in%
               c("Frontal Cortex",
                 "Lumbar Spinal Cord",
                 "Cervical Spinal Cord",
                 "Motor Cortex",
                 "Temporal Cortex",
                 "Cerebellum") )%>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  filter(disease_tissue == T) %>%
  filter(UNC13A_3prime_leaf + UNC13A_5prime_1_leaf >= 1) %>%
  filter(STMN2_cryptic_leaf >= 1) %>%
  dplyr::select(sample,disease_group2,tissue_clean,stmn_2_cryptic_psi_leaf,unc13a_cryptic_leaf_psi) %>%
  melt() %>%
    mutate(tissue_clean = fct_relevel(tissue_clean,"Motor Cortex", "Cervical Spinal Cord",
                                    "Frontal Cortex",
                                    "Lumbar Spinal Cord",
                                    "Temporal Cortex")) %>%
  mutate(variable = ifelse(grepl("stmn_2",variable), "STMN2", "UNC13A")) %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(aes(color = variable),outlier.shape = NA) +
  geom_jitter(aes(fill = variable),pch = 21,height = 0,width = 0.2,size = 3) +
  scale_fill_manual(values = c("#008574","#cd5582")) +
  scale_color_manual(values = c("#008574","#cd5582")) +
  facet_wrap(~tissue_clean,nrow = 3, scales = 'free') +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L))  +
  ggpubr::theme_pubr() +
  theme(text = element_text(size = 24))  +
  xlab("") +
  ylab("CE PSI")  +
  theme(legend.position = 'none')

@@

```{r reproduce_4a_in_rap1gap}
detected_table_for_graph_rap1 = clean_data_table %>%
  filter(!(rs12973192 == "C/G" & rs12608932 == "C/C")) %>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  filter(rap1gap_cryptic_psi > 0) %>%
  filter(stmn_2_cryptic_psi_leaf > 0 ) %>%
  filter(junction_reads_stmn2 >= 30) %>%
  filter((RAP1GAP_noveldonor + RAP1GAP_novelacceptor + RAP1GAP_annotated) >=30) %>% 
  filter(grepl("Cortex",tissue_clean)) %>%
  filter(disease_tissue == T) %>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  dplyr::select(sample,tissue_clean,disease_group2,rs12973192,rs12608932,stmn_2_cryptic_psi_leaf,rap1gap_cryptic_psi) %>%
  mutate(log2_rap1gap = log2((rap1gap_cryptic_psi)/(stmn_2_cryptic_psi_leaf))) %>%
  mutate(tissue_clean = gsub("_","\n",tissue_clean)) %>%
  group_by(rs12973192) %>%
  mutate(n_sample = length(unique(sample))) %>%
  ungroup() %>%
  mutate(detection_name = glue::glue("{rs12973192} \n ( {n_sample} )")) %>%
  mutate(no_cong = ifelse(rs12608932 == "C/C" & rs12973192 == "C/G",8,3)) %>% as.data.table()

@@

detected_table_for_graph_rap1 %>%
  ggplot(aes(y = log2_rap1gap, x = detection_name,fill = detection_name)) +
  geom_boxplot(show.legend = F) +
  geom_jitter(show.legend = F, alpha = 0.8, height = 0,pch = 21) +
  stat_compare_means() +
  stat_compare_means(aes(group = detection_name),comparisons = comparisons,label = "p.signif") +
  ylab("Ratio RAP1GAP CE PSI / STMN2 CE PSI") +
  ylab(expression(paste("Lo", g[2], " fold UNC13A CE PSI / STMN2 CE PSI "))) +
  xlab("UNC13A rs12973192 Genotype") +
  scale_fill_manual(values = c("#88CCEE","#44AA99","#1B7739")) +
  ggpubr::theme_pubr() +
  theme(text = element_text(size = 24)) +
  geom_hline(yintercept = 0,size = 1.5)

@@

>>>>

## Read length affects detection probability in cases

```{r readlength_detection_in_cases}
# library size 
#llumina HiSeq 2500 (125 bp paired end) or an Illumina NovaSeq (100 bp paired end). 
case_control_sequencing <- clean_data_table %>%
  filter(tissue_clean %in% c("Frontal_Cortex", 
                             "Lumbar_Spinal_Cord", 
                             "Cervical_Spinal_Cord", 
                             "Motor_Cortex", 
                             "Temporal_Cortex",
                             "Cerebellum")) %>% 
  filter(sequencing_platform != "")

@@

    p_value = chisq.test(comparison_table) %>%  broom::tidy() %>% pull(p.value)

    print(p_value)

@@

plot_list = list()
for(t in 1:length(lib_comps)){

        groups = c("ALS-TDP", "Control", "ALS \n non-TDP", "FTLD-TDP", "FTLD \n non-TDP")

@@

        ggplot(aes(x = disease_group2,y = rin,fill =disease_group2 )) + 
        geom_boxplot(colour = "black", outlier.colour = NA) +
        geom_point(pch = 21, size = 2, alpha = 0.9, position = position_jitter(width = 0.2, height = 0) ) + 
        facet_wrap(~ tissue_clean,scales = 'free_x') + 
        # ggpubr::stat_compare_means(label = "p.format", label.x.npc = "center", method = "wilcox" ) +
        ggpubr::stat_compare_means(label = "p.signif",
                                   label.x.npc = "center",
                                   method = "wilcox",
                                   comparisons = list(
                                       c("Control","ALS-TDP"),
                                       c("ALS \n non-TDP","ALS-TDP"))) +
        ggpubr::stat_compare_means( label = "p.signif",
                                   label.x.npc = "center", 
                                   method = "wilcox",
                                   comparisons = list(
                                       c("Control","FTLD-TDP"),
                                       c("FTLD-TDP","FTLD \n non-TDP"))) + 
        labs(y = "RIN", x = element_blank()) +
        theme(axis.text.x = element_text(size = 15)) + 
        ggpubr::theme_pubr() + 
        theme(axis.ticks.x = element_blank(),
              axis.text.x = element_blank(),
              legend.position = "none") +
      scale_color_manual(values = force_colors) + 
      scale_fill_manual(values = force_colors)

@@

        groups = c("ALS-TDP", "Control", "ALS \n non-TDP", "FTLD-TDP", "FTLD \n non-TDP")

@@

cell_types %>% 
  left_join(clean_data_table[,.(sample,
                               UNC13A_3prime_leaf,
                               UNC13A_5prime_1_leaf,
                               disease_group2,
                               disease_tissue)], by = 'sample') %>%
  filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  filter(disease_tissue == T) %>% 
  filter(tissue_clean != "Thoracic_Spinal_Cord") %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-")) %>% 
  add_count(disease_group2,tissue_clean) %>% 
  mutate(n = n / 5) %>% 
  mutate(plot_name = glue::glue("{disease_group2} \n {n}")) %>% 
  filter(!is.na(inclusion_reads)) %>% 
  unique() %>% 
  ggplot(aes(x = cell, y = deconv, fill = unc13a_detected)) + 
  geom_point(size = 2,show.legend = F,pch = 21, position = position_jitterdodge(dodge.width=1,jitter.width = 0.3))+
  geom_boxplot(outlier.color = NA) + 
  facet_wrap(~plot_name+tissue_clean,scales = 'free') +
  ggpubr::stat_compare_means(label = "p.signif") +
  ggpubr::stat_compare_means(label = "p.format",vjust = 1.2) +
  labs(title = "Cell type decomposition and UNC13A CE detection ",subtitle = "Wilcox test") + 
  ggpubr::theme_pubr()

@@

```{r onlyshort}

star_maker = function (p.value) 
{
    unclass(symnum(p.value, corr = FALSE, na = FALSE, cutpoints = c(0, 0.0001,
        0.001, 0.01, 0.05, 0.1, 1), symbols = c("****","***", "**", 
        "*", ".", " ")))
}
only_short_plotter = boxplot_table %>%
    filter(tissue_clean %in%
               c("Frontal Cortex",
                 "Lumbar Spinal Cord",
                 "Cervical Spinal Cord",
                 "Motor Cortex",
                 "Temporal Cortex",
                 "Cerebellum") )%>%
    group_by(disease_group2,tissue_clean) %>%
    mutate(n_sample = n_distinct(sample)) %>%
    ungroup() %>%
    mutate(disease_group2 = gsub("Control", " Control",disease_group2)) %>%
    mutate(detection_name = glue::glue("{disease_group2} \n ( {n_sample} )"))  %>%
    mutate(tissue_clean = fct_relevel(tissue_clean, "Cervical Spinal Cord",
                                      "Frontal Cortex",
                                      "Lumbar Spinal Cord",
                                      "Temporal Cortex",
                                      "Motor Cortex"))

only_short_plotter %>% 
  dplyr::select(sample,tissue_clean,inclusion_reads,disease_group2) %>% 
  mutate(disease_group2 = gsub("\n", "",disease_group2)) %>% 
  unique() %>% 
  group_by(tissue_clean) %>% 
  nest() %>%
  mutate(wilcox_res = map(data, ~pairwise.wilcox.test(.x$inclusion_reads,.x$disease_group2) %>% broom::tidy())) %>%
  dplyr::select(-data) %>% 
  unnest() %>% 
  mutate(star_plot = star_maker(p.value)) %>% fwrite("~/Desktop/stars_for_weaverly_fig3b.csv")

only_short_plotter %>% 
    ggplot(aes(x = detection_name,
               y = inclusion_reads * 10^6,
               fill ="red")) +
    geom_boxplot(show.legend = F,position = position_dodge(preserve = "single",width = 1),outlier.colour = NA) +
    geom_point(size = 2,show.legend = F,pch = 21, position = position_jitterdodge(dodge.width=1,jitter.width = 0.3))+
    ggplot2::facet_wrap(vars(tissue_clean),scales = "free_x",nrow = 3) +
    ylab("UNC13A cryptic \n reads per million") +
    theme(text = element_text(size = 12)) +
    xlab("") +
    scale_fill_manual(values = colorblind_pal()(4)[2:4]) +
    theme_bw() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(text = element_text(size = 18))  +
    theme(legend.position = "bottom") + 
    stat_compare_means(comparisons = comps)
```

@@

## Side by side comparison UNC13A & STMN2 CE PSI
```{r}
clean_data_table %>%
    mutate(tissue_clean = gsub("_"," ",tissue_clean)) %>%
    filter(tissue_clean %in%
               c("Frontal Cortex",
                 "Lumbar Spinal Cord",
                 "Cervical Spinal Cord",
                 "Motor Cortex",
                 "Temporal Cortex",
                 "Cerebellum") )%>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  filter(disease_tissue == T) %>%
  filter(UNC13A_3prime_leaf + UNC13A_5prime_1_leaf >= 1) %>%
  filter(STMN2_cryptic_leaf >= 1) %>%
  dplyr::select(sample,disease_group2,tissue_clean,stmn_2_cryptic_psi_leaf,unc13a_cryptic_leaf_psi) %>%
  melt() %>%
    mutate(tissue_clean = fct_relevel(tissue_clean,"Motor Cortex", "Cervical Spinal Cord",
                                    "Frontal Cortex",
                                    "Lumbar Spinal Cord",
                                    "Temporal Cortex")) %>%
  mutate(variable = ifelse(grepl("stmn_2",variable), "STMN2", "UNC13A")) %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(aes(color = variable),outlier.shape = NA) +
  geom_jitter(aes(fill = variable),pch = 21,height = 0,width = 0.2,size = 3) +
  scale_fill_manual(values = c("#008574","#cd5582")) +
  scale_color_manual(values = c("#008574","#cd5582")) +
  facet_wrap(~tissue_clean,nrow = 3, scales = 'free') +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L))  +
  ggpubr::theme_pubr() +
  theme(text = element_text(size = 24))  +
  xlab("") +
  ylab("CE PSI")  +
  theme(legend.position = 'none') + 
  stat_compare_means(label = 'p.signif')

@@

```{r reproduce_4a_in_rap1gap}
detected_table_for_graph_rap1 = clean_data_table %>%
  filter(!(rs12973192 == "C/G" & rs12608932 == "C/C")) %>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  filter(rap1gap_cryptic_psi > 0) %>%
  filter(stmn_2_cryptic_psi_leaf > 0 ) %>%
  filter(junction_reads_stmn2 >= 30) %>%
  filter((RAP1GAP_noveldonor + RAP1GAP_novelacceptor + RAP1GAP_annotated) >=30) %>% 
  filter(grepl("Cortex",tissue_clean)) %>%
  filter(disease_tissue == T) %>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  dplyr::select(sample,tissue_clean,disease_group2,rs12973192,rs12608932,stmn_2_cryptic_psi_leaf,rap1gap_cryptic_psi) %>%
  mutate(log2_rap1gap = log2((rap1gap_cryptic_psi)/(stmn_2_cryptic_psi_leaf))) %>%
  mutate(tissue_clean = gsub("_","\n",tissue_clean)) %>%
  group_by(rs12973192) %>%
  mutate(n_sample = length(unique(sample))) %>%
  ungroup() %>%
  mutate(detection_name = glue::glue("{rs12973192} \n ( {n_sample} )")) %>%
 as.data.table()

@@

rap1_fold_plot = detected_table_for_graph_rap1 %>%
  ggplot(aes(y = log2_rap1gap, x = detection_name,fill = detection_name)) +
  geom_boxplot(show.legend = F) +
  geom_jitter(show.legend = F, alpha = 0.8, height = 0,pch = 21) +
  stat_compare_means(aes(group = detection_name),comparisons = comparisons,label = "p.signif") +
  ylab(expression(paste("Lo", g[2], " fold CE PSI / STMN2 CE PSI "))) +
  xlab("UNC13A rs12973192 Genotype") +
  scale_fill_manual(values = c("#88CCEE","#44AA99","#1B7739")) +
  ggpubr::theme_pubr() +
  theme(text = element_text(size = 24)) +
  geom_hline(yintercept = 0,size = 1.5) + 
  ggtitle("RAP1GAP")

@@

```{r reproduce_4a_in_pfkp}
detected_table_for_graph_pfkp = clean_data_table %>%
  filter(!(rs12973192 == "C/G" & rs12608932 == "C/C")) %>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  filter(pfkp_cryptic_psi > 0) %>%
  filter(stmn_2_cryptic_psi_leaf > 0 ) %>%
  filter(junction_reads_stmn2 >= 30) %>%
    filter((PFKP_novelacceptor + PFKP_noveldonor + PFKP_annotated) >=30) %>% 
  filter(grepl("Cortex",tissue_clean)) %>%
  filter(disease_tissue == T) %>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  dplyr::select(sample,tissue_clean,disease_group2,rs12973192,rs12608932,stmn_2_cryptic_psi_leaf,pfkp_cryptic_psi) %>%
  mutate(log2_pfkp = log2((pfkp_cryptic_psi)/(stmn_2_cryptic_psi_leaf))) %>%
  mutate(tissue_clean = gsub("_","\n",tissue_clean)) %>%
  group_by(rs12973192) %>%
  mutate(n_sample = length(unique(sample))) %>%
  ungroup() %>%
  mutate(detection_name = glue::glue("{rs12973192} \n ( {n_sample} )")) %>%
  mutate(no_cong = ifelse(rs12608932 == "C/C" & rs12973192 == "C/G",8,3)) %>% as.data.table()

comparisons = combn(unique(detected_table_for_graph_pfkp$detection_name),2,simplify = F)

pfkp_fold_plot = detected_table_for_graph_pfkp %>%
  ggplot(aes(y = log2_pfkp, x = detection_name,fill = detection_name)) +
  geom_boxplot(show.legend = F) +
  geom_jitter(show.legend = F, alpha = 0.8, height = 0,pch = 21) +
  stat_compare_means(aes(group = detection_name),comparisons = comparisons,label = "p.signif") +
  xlab("UNC13A rs12973192 Genotype") +
  scale_fill_manual(values = c("#88CCEE","#44AA99","#1B7739")) +
  ggpubr::theme_pubr() +
  theme(text = element_text(size = 24)) +
  geom_hline(yintercept = 0,size = 1.5) + 
  ggtitle("PFKP") + 
  ylab(element_blank())

cowplot::plot_grid(rap1_fold_plot,pfkp_fold_plot)
